---
title: "Gov 50 Final Project"
author: "Nicole Fintel"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to public opinion and the death penalty. I'm interested in learning more about how public opinion did or didn't influence US states' decisions whether to abolish the death penalty. Similarly, I'm curious how public opinion on capital punishment within a state changed after that state abolished the death penalty.

## Proposal (Milestone 2)

Do state supreme court rulings on capital punishment lead to increased polarization around the issue of capital punishment? 

For my final project, I plan to examine whether polarization around the issue of capital punishment is driven (or ameliorated) by state supreme court rulings on issues related to the death penalty. I will focus specifically on those court cases that involve either the abolition, retention, or reinstatement of capital punishment in a certain state or cases pertaining to the legality of certain methods/modes of execution in a state (i.e., constitutional regulation of the death penalty). I hypothesize that polarization around the issue of capital punishment within a given state will increase after a state supreme court rules on the death penalty. I further hypothesize that these effects will be strongest in cases where a court abolishes the death penalty. Gerald Rosenberg (1991) finds that the U.S. Supreme Court has little power to bring about lasting social change if cultural, political, and economic conditions are not amenable to said social change. Instead, rulings on contentious social issues may spark countermovements that ultimately drive progress in the opposite direction. Similarly, Gordon Silverstein (2004) argues that “Court rulings can goad social movements affirmatively [...] or negatively, as groups organize in resistance to, or rejection of, a court ruling.” In analyzing 50 years of state- and national-level Democratic and Republican party platforms, my own previous research for Gov 97 and Gov 94 has found that Supreme Court rulings on contentious social issues appear to lead to an increase in polarization at both the national- and state- level, driven primarily by conservative countermovements. A number of scholars (Barber and McCarty, 2013; Hacker and Pierson 2020; Thomsen, 2014; etc.) have shown that the Republican Party has played an outsized role in driving this partisan polarization. That said, much less is known about the effects of state supreme court rulings, and very little is known about polarization surrounding capital punishment in particular. 

The explanatory variable of interest will be whether a state supreme court issued a ruling on the legality of capital punishment in a given year. I will measure this by using online archives to discover when states issued these rulings. Years with a relevant court ruling will be coded as a 1 and years without will be coded as a 0.

The outcome variable will be polarization, which I will measure as the extent to which public opinion about the death penalty among Democrats and Republicans differs. For instance, if 30% of Democrats and 70% of Republicans supported capital punishment in a given year, I’d code this variable as =40, because 40 points separate D and R public opinion.

If I observe an increase in the divergence of public support for the death penalty among Democrats and Republicans (ie., an increase in polarization)—that sharply outpaces national trends—in a state after a relevant court ruling, then this will provide support for my hypothesis. Conversely, if I observe no discernible change or a decrease in polarization after a relevant court ruling, then that would provide evidence against my hypothesis.

I plan to use opinion poll data from the General Social Survey (GSS) and/or the Gallup Poll. GSS has data from 1974-2022 for the question “Do you favor or oppose the death penalty for persons convicted of murder?” and Gallup has data from 1953-2022 for the question, “Are you in favor of the death penalty for a person convicted of murder?” I will be able to select data for various states in the years before, during, and after that state’s supreme court issued a ruling on the death penalty and compare it to national trends. 


## Milestone 3

Note: I will likely be altering my intial proposal a bit due to difficulties obtaining data for individual states. I plan to speak to a TF about this in more detail soon. In the meantime, I've made two other graphs that illustrate the distribution of executions in the US over time, two graphs that show how American public opinion on capital punishment has changed over time, and one graph that plots the relationship between public support for capital punishment and the number of executions.

Sources: https://gssdataexplorer.norc.org/trends for public opinion data,  https://bjs.ojp.gov/topics/corrections/capital-punishment for execution data through 2020, and Amnesty International for data on executions in 2021 and 2022.

```{r}
library(tidyverse)
library(dplyr)

cappun <- read_csv("cappun3.csv")

cappun_plot_1 <- ggplot(data = cappun, mapping = aes(x = Year, y = Number_Executions)) + geom_line(color = "red") +
  labs(x = "Year", y = "Number of Executions", title = "Number of Executions in the United States per Year", caption = "Sources: US Department of Justice & Amnesty International")

cappun_plot_2 <- ggplot(data = cappun,
mapping = aes(x = Year, y = Number_Executions)) +
geom_point() +
geom_smooth(color = "red") +
  labs(x = "Year", y = "Number of Executions", title = "Number of Executions in the United States per Year", caption = "Sources: US Department of Justice & Amnesty International")

cappun_plot_1
cappun_plot_2
```

```{r}
cappun_plot_3 <- ggplot(data = cappun, mapping = aes(x = Year, y = Percent_Support)) + geom_line() +
  labs(x = "Year", y = "Percent in Favor", title = "% of Americans in Favor of Capital Punishment for Homicide by Year", caption = "Source: General Social Survey")

cappun_plot_4 <- ggplot(data = cappun,
mapping = aes(x = Year, y = Percent_Support)) +
geom_point() +
geom_smooth(color = "black") +
    labs(x = "Year", y = "Percent in Favor", title = "% of Americans in Favor of Capital Punishment for Homicide by Year", caption = "Source: General Social Survey")

cappun_plot_3
cappun_plot_4
```


```{r}
cappun_plot_5 <- ggplot(data = cappun,
mapping = aes(x = Percent_Support, y = Number_Executions)) +
geom_point() +
geom_smooth(color = "Blue") +
  labs(x = "Percent in Favor", y = "Number of Executions", title = "Number of Executions vs. Support for Capital Punishment", caption = "Sources: General Social Survey, US Department of Justice, and Amnesty International")

cappun_plot_5
cor(cappun$Percent_Support, cappun$Number_Executions)
```

```{r}
library(tidyverse)
library(dplyr)
cappun3 <- read_csv("cappun3.csv")
cappun3

cappun_plot_6 <- ggplot(data = cappun3,
mapping = aes(y = Percent_Support, x = Number_Exonerations)) +
geom_point() +
geom_smooth(color = "Blue")

cappun_plot_6
cor(cappun3$Number_Exonerations, cappun3$Percent_Support)

cappun_plot_7 <- ggplot(data = cappun3,
mapping = aes(y = Number_Executions, x = Number_Exonerations)) +
geom_point() +
geom_smooth(color = "Blue")
cappun_plot_7

cor(cappun3$Number_Exonerations, cappun3$Number_Executions)
```

## Milestone 4 

```{r}
library(tidyverse)
library(dplyr)

cappun_table <- read_csv("cappun3.csv")
  lm(Number_Executions ~ Number_Exonerations, data = cappun3)

library(broom)
coef(cappun_table)
  
  knitr::kable(cappun_table)
```
## Analysis
The intercept is 13.434, which means that when the number of exonerations is zero, then the average number of executions will be approximately 13.4 (holding all other variables constant). The coefficient for the number of exonerations in 3.541, which means that, holding all other variables constant, a one-unit increase in the number of exonerations (i.e., each additional exoneration) is associated with a 3.5-unit increase in executions (i.e., 3.5 additional executions). This finding was a bit surprising to me, as I had hypothesized that exonerations would work to delegitimize the institution of capital punishment by pointing out its flaws. Instead, this data seems to suggest it may actually legitimize capital punishment by showing that mistakes can be amended. 



## Final Project

```{r}
library(tidyverse)
library(dplyr)
library(broom)
library(infer)
library(modelsummary)

deathpenalty <- read_csv("deathpenalty.csv")
knitr::kable(deathpenalty)

```


```{r}
dp_plot_1 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Number_Executions)) + 
  geom_line(color = "red") + 
  labs(x = "Year", 
       y = "Number of Executions", 
       title = "Number of Executions in the United States per Year", 
       caption = "Sources: US Department of Justice")

dp_plot_2 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Number_Executions)) + 
  geom_point() + geom_smooth(color = "red") + 
  labs(x = "Year", 
       y = "Number of Executions", 
       title = "Number of Executions in the United States per Year", 
       caption = "Sources: US Department of Justice")

dp_plot_3 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Number_Exonerations)) + 
  geom_line(color = "forestgreen") + 
  labs(x = "Year", 
       y = "Number of Exonerations", 
       title = "Number of Exonerations in the United States per Year", 
       caption = "Sources: US Department of Justice")

dp_plot_4 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Number_Exonerations)) + 
  geom_point() + 
  geom_smooth(color = "forestgreen") + 
  labs(x = "Year", 
       y = "Number of Exonerations", 
       title = "Number of Exonerations in the United States per Year", 
       caption = "Sources: US Department of Justice")

dp_plot_5 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Percent_Support)) + 
  geom_line() +
  labs(x = "Year", 
       y = "Percent in Favor", 
       title = "% of Americans in Favor of Capital Punishment for Homicide by Year", 
       caption = "Source: General Social Survey") 
  

dp_plot_6 <- ggplot(data = deathpenalty,  
  mapping = aes(x = Year, y = Percent_Support)) + 
  geom_point() + 
  geom_smooth(color = "black") + 
  labs(x = "Year", 
       y = "Percent in Favor", 
       title = "% of Americans in Favor of Capital Punishment for Homicide by Year", 
       caption = "Source: General Social Survey")



dp_plot_1
dp_plot_2
dp_plot_3
dp_plot_4
dp_plot_5
dp_plot_6
```



```{r}
dp_plot_7 <- ggplot(data = deathpenalty,
mapping = aes(x = Percent_Support, y = Number_Executions)) +
geom_point() +
geom_smooth(color = "Blue") +
  labs(x = "Percent in Favor", y = "Number of Executions", title = "Number of Executions vs. Support for Capital Punishment", caption = "Sources: General Social Survey & US Department of Justice")




dp_plot_8 <- ggplot(data = deathpenalty,
mapping = aes(x = Percent_Support, y = Number_Executions_Lag)) +
geom_point() +
geom_smooth(color = "Blue") +
  labs(x = "Percent in Favor, Year Prior", y = "Number of Executions", title = "Number of Executions vs. Support for Capital Punishment (Lag)", caption = "Sources: General Social Survey & US Department of Justice")

dp_plot_7
dp_plot_8
cor(deathpenalty$Percent_Support, deathpenalty$Number_Executions)
cor(deathpenalty$Percent_Support, deathpenalty$Number_Executions_Lag)

```



```{r}
lm(Number_Executions ~ Percent_Support, data = deathpenalty)
lm(Number_Executions_Lag ~ Percent_Support, data = deathpenalty)
lm(Number_Executions ~ Percent_Support_Lag, data = deathpenalty)

lm(Percent_Support ~ Homicide_Rate, data = deathpenalty)


```




```{r}

lm_support_1 <- lm(Percent_Support ~ Homicide_Rate,
              data = deathpenalty) 

summary(lm_support_1)
modelsummary::modelsummary(lm_support_1,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))




lm_support_2 <- lm(Percent_Support ~ Fin_Dissatisfaction,
              data = deathpenalty) 

summary(lm_support_2)
modelsummary::modelsummary(lm_support_2,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))





lm_support_3 <- lm(Percent_Support ~ Homicide_Rate + Fin_Dissatisfaction,
              data = deathpenalty) 

summary(lm_support_3)
modelsummary::modelsummary(lm_support_3,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```


```{r}
lm_executions_1 <- lm(Number_Executions ~ Homicide_Rate,
              data = deathpenalty) 

summary(lm_executions_1)
modelsummary::modelsummary(lm_executions_1,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))


```
```{r}
lm_executions_2 <- lm(Number_Executions ~ Percent_Support,
              data = deathpenalty) 

summary(lm_executions_2)
modelsummary::modelsummary(lm_executions_2,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```



```{r}
lm_executions_3 <- lm(Number_Executions ~ Homicide_Rate + Percent_Support,
              data = deathpenalty) 

summary(lm_executions_3)
modelsummary::modelsummary(lm_executions_3,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```


```{r}
lm_executions_4 <- lm(Number_Executions ~ Number_Exonerations,
              data = deathpenalty) 

summary(lm_executions_4)
modelsummary::modelsummary(lm_executions_4,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```


```{r}
lm_executions_5 <- lm(Number_Executions ~ Number_Exonerations_Lag,
              data = deathpenalty) 

summary(lm_executions_5)
modelsummary::modelsummary(lm_executions_5,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```


```{r}



```

