---
title: "Gov 50 Final Project"
author: "Nicole Fintel"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---


## Introduction
In Furman v. Georgia (1972), the United States Supreme Court ruled that capital punishment, in its current form, was being applied in an arbitrary, capricious, and racially biased manner. Such application of capital punishment, according to the Court, constituted a "cruel and unusual punishment," and therefore violated the 8th and 14th amendments of the US Constitution. In effect, the Furman ruling placed an indefinite moratorium on executions until states and the federal government could alter their capital statutes in a manner that no longer violated the Constitution. Four years later, in Gregg v. Georgia (1976), the Court ruled that Georgia that a handful of states' revised statutes conformed to the standards required by Furman, thus lifting the moratorium on executions and ushering in what scholars has dubbed "the modern era of the American death penalty." The reinstatement of capital punishment in the Untied States was a historic anomaly, as the majority of America's peer nations during this time period were moving to permanently abolish the death penalty. Indeed, today, the Untied States is the world's only "western" democracy to retain its death penalty and to actively execute its citizens. 

Both today and in the years preceding Furman v. Georgia, death penalty abolitionists and anti-death penalty activists have pursued a "moratorium strategy," where they attempt to get state governments and the federal governments to place a moratorium on executions, or at least minimize the annual number of executions. As explained by ___() lowering execution rates is often the first step toward permanent abolition of capital punishment, as was the case in ___. Some scholars believe that that, by substantially lowering the number of annual executions, permanent death penalty abolition in the United States will become more feasible. However, there is currently a dearth of research on the factors/conditions that impact the number of executions the United States carries out in a given year. My project aims to begin to fill this gap in research. This research is important because it may help activists better focus their efforts on those factors or conditions that decrease the number of annual executions.

To this end, I ask the following research question:



The corpus of social science literature on capital punishment emphasizes that, in most democratic nations, public opinion plays a limited role in governments' decision to abolish the death penalty. However, these scholars note that it is unwise the assume these trends will hold true in the Unites States, as American democracy has a number of idiosyncratic features that change the calculus. These features include (but are not limited to) popularly elected prosecutors and judges, the frequent use of referenda and ballot initiatives,  Ultimately, ____. Nonetheless, the American states that have previously abolished the death penalty managed to do so in spite of public opinion, but because of it. Guided by this literature, I have formulated my first hypothesis:

H1: Public support for capital punishment has no impact on the number of executions.


Additionally, ___ [crime rate] As such, I've developed a second hypothesis:

H2: The number of executions increases as the homicide rate increases.

Finally, Carol and Jordan Steiker () explain that ____, the US became the "first and only one of its peer nations to move not from formal retention of the death penalty to abolition, but rather from retention to regulation." This procedural regulation, according to Steiker & Steiker () has had the “effect of legitimating the use of capital punishment as a penal sanction in the eyes of actors within the criminal justice system and the public at large." In other words, in our era of ____ exonerations. Guided by the research of Steiker & Steiker, I have formulated a third (and final) hypothesis:

H3: The number of executions increases as the number of exonerations increases.


## Data & Research Design




I then loaded this file into RStudio, named it "deathpenalty," and presented my dataset in a table. 

```{r}
library(tidyverse)
library(dplyr)
library(broom)
library(infer)
library(modelsummary)

deathpenalty <- read_csv("deathpenalty.csv")
knitr::kable(deathpenalty)

```

To visualize key trends in my dataset, I first plotted the number of executions per year, both as a line graph and as a scatter plot with a smoothed trend line.

```{r}
dp_plot_1 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Number_Executions)) + 
  geom_line(color = "red") + 
  labs(x = "Year", 
       y = "Number of Executions", 
       title = "Number of Executions in the United States per Year", 
       caption = "Source: US Department of Justice")

dp_plot_2 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Number_Executions)) + 
  geom_point() + geom_smooth(color = "red") + 
  labs(x = "Year", 
       y = "Number of Executions", 
       title = "Number of Executions in the United States per Year", 
       caption = "Source: US Department of Justice")

dp_plot_1
dp_plot_2
```

Then, I plotted the key explanatory variable for each of my three hypotheses over time as both a line graph and a scatter plot with a smoothed trend line. These variables are percentage of Americans in favor of capital punishment, national homicide rate, and number of exonerations, respectively.

```{r}
dp_plot_3 <- ggplot(data = deathpenalty[complete.cases(deathpenalty$Percent_Support),],
  mapping = aes(x = Year, y = Percent_Support)) + 
  geom_line() +
  labs(x = "Year", 
       y = "Percent in Favor", 
       title = "% of Americans in Favor of Capital Punishment for Homicide by Year", 
       caption = "Source: General Social Survey") 
  

dp_plot_4 <- ggplot(data = deathpenalty,  
  mapping = aes(x = Year, y = Percent_Support)) + 
  geom_point() + 
  geom_smooth(color = "black") + 
  labs(x = "Year", 
       y = "Percent in Favor", 
       title = "% of Americans in Favor of Capital Punishment for Homicide by Year", 
       caption = "Source: General Social Survey")

dp_plot_5 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Homicide_Rate)) + 
  geom_line(color = "darkorchid4") + 
  labs(x = "Year", 
       y = "Homicide Rate", 
       title = "Homicide Rate in the United States per Year", 
       caption = "Source: Federal Bureau of Investigations")

dp_plot_6 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Homicide_Rate)) + 
  geom_point() +
  geom_smooth(color = "darkorchid4") + 
  labs(x = "Year", 
       y = "Homicide Rate", 
       title = "Homicide Rate in the United States per Year", 
       caption = "Source: Federal Bureau of Investigations")

dp_plot_7 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Number_Exonerations)) + 
  geom_line(color = "forestgreen") + 
  labs(x = "Year", 
       y = "Number of Exonerations", 
       title = "Number of Exonerations in the United States per Year", 
       caption = "Source: US Department of Justice")

dp_plot_8 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Number_Exonerations)) + 
  geom_point() + 
  geom_smooth(color = "forestgreen") + 
  labs(x = "Year", 
       y = "Number of Exonerations", 
       title = "Number of Exonerations in the United States per Year", 
       caption = "Source: US Department of Justice")

dp_plot_3
dp_plot_4
dp_plot_5
dp_plot_6
dp_plot_7
dp_plot_8
```


##Results




```{r}
dp_plot_7 <- ggplot(data = deathpenalty,
mapping = aes(x = Percent_Support, y = Number_Executions)) +
geom_point() +
geom_smooth(color = "Blue") +
  labs(x = "Percent in Favor", y = "Number of Executions", title = "Number of Executions vs. Support for Capital Punishment", caption = "Sources: General Social Survey & US Department of Justice")




dp_plot_8 <- ggplot(data = deathpenalty,
mapping = aes(x = Percent_Support, y = Number_Executions_Lag)) +
geom_point() +
geom_smooth(color = "Blue") +
  labs(x = "Percent in Favor, Year Prior", y = "Number of Executions", title = "Number of Executions vs. Support for Capital Punishment (Lag)", caption = "Sources: General Social Survey & US Department of Justice")

dp_plot_7
dp_plot_8
cor(deathpenalty$Percent_Support, deathpenalty$Number_Executions, use = 'complete.obs')
cor(deathpenalty$Percent_Support, deathpenalty$Number_Executions_Lag, use = 'complete.obs')

```



```{r}
lm(Number_Executions ~ Percent_Support, data = deathpenalty)
lm(Number_Executions_Lag ~ Percent_Support, data = deathpenalty)
lm(Number_Executions ~ Percent_Support_Lag, data = deathpenalty)

lm(Percent_Support ~ Homicide_Rate, data = deathpenalty)


```





```{r}
lm_executions_1 <- lm(Number_Executions ~ Homicide_Rate,
              data = deathpenalty) 

summary(lm_executions_1)
modelsummary::modelsummary(list(Regression = lm_executions_1),
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))


```
```{r}
lm_executions_2 <- lm(Number_Executions ~ Percent_Support,
              data = deathpenalty) 

summary(lm_executions_2)
modelsummary::modelsummary(list(Regression = lm_executions_2),
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```



```{r}
lm_executions_3 <- lm(Number_Executions ~ Homicide_Rate + Percent_Support,
              data = deathpenalty) 

summary(lm_executions_3)
modelsummary::modelsummary(list(Regression = lm_executions_3),
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```


```{r}
lm_executions_4 <- lm(Number_Executions ~ Number_Exonerations,
              data = deathpenalty) 

summary(lm_executions_4)
modelsummary::modelsummary(list(Regression = lm_executions_4),
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```


```{r}
lm_executions_5 <- lm(Number_Executions ~ Number_Exonerations_Lag,
              data = deathpenalty) 

summary(lm_executions_5)
modelsummary::modelsummary(list(Regression = lm_executions_5),
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))

```

For a bit of secondary analysis, I decided to investigate potential factors or conditions that impact public support for capital punishment. First, to explore the relationship between the homicide rate and public support for the death penalty, I ran a regression using "Percent_Support" as the response variable and "Homicide_Rate" as the explanatory variable and plotted these two variables on a scatter plot with a smoothed trend line. 

```{r}
lm_support_1 <- lm(Percent_Support ~ Homicide_Rate,
              data = deathpenalty) 

summary(lm_support_1)
modelsummary::modelsummary(list(Regression = lm_support_1),
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))

support_plot_1 <- ggplot(data = deathpenalty,
  mapping = aes(x = Homicide_Rate, y = Percent_Support)) + 
  geom_point() + geom_smooth(color = "black") + 
  labs(x = "Homicide Rate", 
       y = "Percent in Favor", 
       title = "Percent in Favor vs. Homicide Rate", 
       caption = "Sources: General Social Survey & US Department of Justice")

support_plot_1
```
According to this regression, the intercept is 56.884, which means that — all else being constant — when the homicide rate is 0,  public support for capital punishment is 56.884%. The coefficient for homicide rate is 1.311, which means that — all else being held constant — for each unit (percentage point) the homicide rate increases, public support for capital punishment increases by 1.311 units (percentage points). This relationship is statistically significant, as the p-value is under 0.05.


Then, to explore the relationship between the rate of financial dissatisfaction and public support for the death penalty, I ran a regression using "Percent_Support" as the response variable and "Fin_Dissatisfaction" as the explanatory variable. I also plotted these two variables on a scatter plot with a smoothed trend line. 

```{r}
lm_support_2 <- lm(Percent_Support ~ Fin_Dissatisfaction,
              data = deathpenalty) 

summary(lm_support_2)
modelsummary::modelsummary(list(Regression = lm_support_2),
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))

support_plot_2 <- ggplot(data = deathpenalty,
  mapping = aes(x = Fin_Dissatisfaction, y = Percent_Support)) + 
  geom_point() + geom_smooth(color = "black") + 
  labs(x = "Financial Dissatisfaction", 
       y = "Percent in Favor", 
       title = "Percent in Favor vs. Financial Dissatisfaction", 
       caption = "Sources: General Social Survey & US Department of Justice")

support_plot_2
```
According to this regression, the intercept is 61.167, which means that — all else being constant — when the homicide rate is 0,  public support for capital punishment is 61.167%. The coefficient for homicide rate is 0.223, which means that — all else being held constant — for each unit (FIGURE OUT UNIT) financial dissatisfaction increases, public support for capital punishment increases by 0.223 units (percentage points). However, this relationship is NOT statistically significant, as the p-value is well above 0.05.


Finally, I ran a multiple regression using Percent_Support as the response variable and "Homicide_Rate" and "Fin_Dissatisfaction" as the explanatory variables.

```{r}
lm_support_3 <- lm(Percent_Support ~ Homicide_Rate + Fin_Dissatisfaction,
              data = deathpenalty) 

summary(lm_support_3)
modelsummary::modelsummary(list(Regression = lm_support_3),
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```
In this multiple regression, the relationship between the homicide rate and public support remains statistically significant and the relationship between the rate of financial dissatisfaction and public support remains not statistically significant. However, 


