---
title: "Gov 50 Final Project"
author: "Nicole Fintel"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Milestone 4 

```{r}
library(tidyverse)
library(dplyr)

cappun_table <- read_csv("cappun3.csv")
  lm(Number_Executions ~ Number_Exonerations, data = cappun3)

library(broom)
coef(cappun_table)
  
  knitr::kable(cappun_table)
```
## Analysis
The intercept is 13.434, which means that when the number of exonerations is zero, then the average number of executions will be approximately 13.4 (holding all other variables constant). The coefficient for the number of exonerations in 3.541, which means that, holding all other variables constant, a one-unit increase in the number of exonerations (i.e., each additional exoneration) is associated with a 3.5-unit increase in executions (i.e., 3.5 additional executions). 



## Introduction
In Furman v. Georgia (1972), the United States Supreme Court ruled that capital punishment, in its current form, was being applied in an arbitrary, capricious, and racially biased manner. Such application of capital punishment, according to the Court, constituted a "cruel and unusual punishment," and therefore violated the 8th and 14th amendments of the US Constitution. In effect, the Furman ruling placed an indefinite moratorium on executions until states and the federal government could alter their capital statutes in a manner that no longer violated the Constitution. Four years later, in Gregg v. Georgia (1976), the Court ruled that Georgia that a handful of states' revised statutes conformed to the standards required by Furman, thus lifting the moratorium on executions and ushering in what scholars has dubbed "the modern era of the American death penalty." The reinstatement of capital punishment in the Untied States was a historic anomaly, as the majority of America's peer nations during this time period were moving to permanently abolish the death penalty. Indeed, today, the Untied States is the world's only "western" democracy to retain its death penalty and to actively execute its citizens. 

Both today and in the years preceding Furman v. Georgia, death penalty abolitionists and anti-death penalty activists have pursued a "moratorium strategy," where they attempt to get state governments and the federal governments to place a moratorium on executions, or at least minimize the annual number of executions. As explained by ___() lowering execution rates is often the first step toward permanent abolition of capital punishment, as was the case in ___. Some scholars believe that that, by substantially lowering the number of annual executions, permanent death penalty abolition in the United States will become more feasible. However, there is currently a dearth of research on the factors/conditions that impact the number of executions the United States carries out in a given year. My project aims to begin to fill this gap in research. This research is important because it may help activists better focus their efforts on those factors or conditions that decrease the number of annual executions.

To this end, I ask the following research question:



The corpus of social science literature on capital punishment emphasizes that, in most democratic nations, public opinion plays a limited role in governments' decision to abolish the death penalty. However, these scholars note that it is unwise the assume these trends will hold true in the Unites States, as American democracy has a number of idiosyncratic features that change the calculus. These features include (but are not limited to) popularly elected prosecutors and judges, the frequent use of referenda and ballot initiatives,  Ultimately, ____. Nonetheless, the American states that have previously abolished the death penalty managed to do so in spite of public opinion, but because of it. Guided by this literature, I have formulated my first hypothesis:

H1: Public support for capital punishment has no impact on the number of executions.


Additionally, ___ [crime rate] As such, I've developed a second hypothesis:

H2: The number of executions increases as the homicide rate increases.

Finally, Carol and Jordan Steiker () explain that ____, the US became the "first and only one of its peer nations to move not from formal retention of the death penalty to abolition, but rather from retention to regulation." This procedural regulation, according to Steiker & Steiker () has had the “effect of legitimating the use of capital punishment as a penal sanction in the eyes of actors within the criminal justice system and the public at large." In other words, in our era of ____ exonerations. Guided by the research of Steiker & Steiker, I have formulated a third (and final) hypothesis:

H3: The number of executions increases as the number of exonerations increases.


## Data


```{r}
library(tidyverse)
library(dplyr)
library(broom)
library(infer)
library(modelsummary)

deathpenalty <- read_csv("deathpenalty.csv")
knitr::kable(deathpenalty)

```


```{r}
dp_plot_1 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Number_Executions)) + 
  geom_line(color = "red") + 
  labs(x = "Year", 
       y = "Number of Executions", 
       title = "Number of Executions in the United States per Year", 
       caption = "Sources: US Department of Justice")

dp_plot_2 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Number_Executions)) + 
  geom_point() + geom_smooth(color = "red") + 
  labs(x = "Year", 
       y = "Number of Executions", 
       title = "Number of Executions in the United States per Year", 
       caption = "Sources: US Department of Justice")

dp_plot_3 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Number_Exonerations)) + 
  geom_line(color = "forestgreen") + 
  labs(x = "Year", 
       y = "Number of Exonerations", 
       title = "Number of Exonerations in the United States per Year", 
       caption = "Sources: US Department of Justice")

dp_plot_4 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Number_Exonerations)) + 
  geom_point() + 
  geom_smooth(color = "forestgreen") + 
  labs(x = "Year", 
       y = "Number of Exonerations", 
       title = "Number of Exonerations in the United States per Year", 
       caption = "Sources: US Department of Justice")

dp_plot_5 <- ggplot(data = deathpenalty[complete.cases(deathpenalty$Percent_Support),],
  mapping = aes(x = Year, y = Percent_Support)) + 
  geom_line() +
  labs(x = "Year", 
       y = "Percent in Favor", 
       title = "% of Americans in Favor of Capital Punishment for Homicide by Year", 
       caption = "Source: General Social Survey") 
  

dp_plot_6 <- ggplot(data = deathpenalty,  
  mapping = aes(x = Year, y = Percent_Support)) + 
  geom_point() + 
  geom_smooth(color = "black") + 
  labs(x = "Year", 
       y = "Percent in Favor", 
       title = "% of Americans in Favor of Capital Punishment for Homicide by Year", 
       caption = "Source: General Social Survey")



dp_plot_1
dp_plot_2
dp_plot_3
dp_plot_4
dp_plot_5
dp_plot_6
```



```{r}
dp_plot_7 <- ggplot(data = deathpenalty,
mapping = aes(x = Percent_Support, y = Number_Executions)) +
geom_point() +
geom_smooth(color = "Blue") +
  labs(x = "Percent in Favor", y = "Number of Executions", title = "Number of Executions vs. Support for Capital Punishment", caption = "Sources: General Social Survey & US Department of Justice")




dp_plot_8 <- ggplot(data = deathpenalty,
mapping = aes(x = Percent_Support, y = Number_Executions_Lag)) +
geom_point() +
geom_smooth(color = "Blue") +
  labs(x = "Percent in Favor, Year Prior", y = "Number of Executions", title = "Number of Executions vs. Support for Capital Punishment (Lag)", caption = "Sources: General Social Survey & US Department of Justice")

dp_plot_7
dp_plot_8
cor(deathpenalty$Percent_Support, deathpenalty$Number_Executions, use = 'complete.obs')
cor(deathpenalty$Percent_Support, deathpenalty$Number_Executions_Lag, use = 'complete.obs')

```



```{r}
lm(Number_Executions ~ Percent_Support, data = deathpenalty)
lm(Number_Executions_Lag ~ Percent_Support, data = deathpenalty)
lm(Number_Executions ~ Percent_Support_Lag, data = deathpenalty)

lm(Percent_Support ~ Homicide_Rate, data = deathpenalty)


```




```{r}

lm_support_1 <- lm(Percent_Support ~ Homicide_Rate,
              data = deathpenalty) 

summary(lm_support_1)
modelsummary::modelsummary(lm_support_1,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))




lm_support_2 <- lm(Percent_Support ~ Fin_Dissatisfaction,
              data = deathpenalty) 

summary(lm_support_2)
modelsummary::modelsummary(lm_support_2,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))





lm_support_3 <- lm(Percent_Support ~ Homicide_Rate + Fin_Dissatisfaction,
              data = deathpenalty) 

summary(lm_support_3)
modelsummary::modelsummary(lm_support_3,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```


```{r}
lm_executions_1 <- lm(Number_Executions ~ Homicide_Rate,
              data = deathpenalty) 

summary(lm_executions_1)
modelsummary::modelsummary(lm_executions_1,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))


```
```{r}
lm_executions_2 <- lm(Number_Executions ~ Percent_Support,
              data = deathpenalty) 

summary(lm_executions_2)
modelsummary::modelsummary(lm_executions_2,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```



```{r}
lm_executions_3 <- lm(Number_Executions ~ Homicide_Rate + Percent_Support,
              data = deathpenalty) 

summary(lm_executions_3)
modelsummary::modelsummary(lm_executions_3,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```


```{r}
lm_executions_4 <- lm(Number_Executions ~ Number_Exonerations,
              data = deathpenalty) 

summary(lm_executions_4)
modelsummary::modelsummary(lm_executions_4,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```


```{r}
lm_executions_5 <- lm(Number_Executions ~ Number_Exonerations_Lag,
              data = deathpenalty) 

summary(lm_executions_5)
modelsummary::modelsummary(lm_executions_5,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```


```{r}



```

