---
title: "Gov 50 Final Project"
author: "Nicole Fintel"
description: "Exploring the impacts of public opinion, homicide rates, and death row exonerations on the number of annual executions in the United States"
output:
  distill::distill_article:
    self_contained: false
---


## Introduction
In Furman v. Georgia (1972), the United States Supreme Court ruled that capital punishment, in its current form, was being applied in an arbitrary, capricious, and racially biased manner. Such application of capital punishment, according to the Court, constituted a "cruel and unusual punishment," and therefore violated the 8th and 14th amendments of the US Constitution. In effect, the Furman ruling placed an indefinite moratorium on executions until states and the federal government could alter their capital statutes in a manner that no longer violated the Constitution. Four years later, in Gregg v. Georgia (1976), the Court ruled that Georgia that a handful of states' revised statutes conformed to the standards required by Furman, thus lifting the moratorium on executions and ushering in what scholars has dubbed "the modern era of the American death penalty" (Killingley, 2008). The reinstatement of capital punishment in the Untied States was a historic anomaly, as the majority of America's peer nations during this time period were moving to permanently abolish the death penalty (Garland, 2010). Indeed, today, the Untied States is the world's only "western" democracy to retain its death penalty and to actively execute its citizens (Steiker & Steiker, 2016). 

Both today and in the years preceding Furman v. Georgia, anti-death penalty activists have pursued a "moratorium strategy," whereby they attempt to institute moratoriums on executions in all retentionist jurisdictions, or—if full moratoriums are not possible—to substantially lower the number of executions carried out (Zimring & Hawkins, 1986). From a historical/comparative perspective, lowering execution rates is often the first step toward achieving permanent abolition of capital punishment, as was the case in Sweden, The Netherlands, Switzerland, and Denmark (Zimring & Hawkins, 1986). Some scholars believe that that—by substantially lowering the number of annual executions—permanent abolition in the United States will become more feasible. However, there is currently relatively little literature/research on the factors or conditions that impact the number of executions carried out each year in the US. My project aims to begin to fill this gap. This research is important because it may help activists better focus their efforts on those factors/conditions that decrease the number of annual executions. It will also help scholars better understand the types of pressures to which governments and lawmakers responsive/receptive.

To this end, I ask the following research question: **Does public support for capital punishment, the homicide rate, or the number of death row exonerations impact the number of executions carried out in the United States?**

The corpus of social science literature on capital punishment emphasizes that, in most democratic nations, public opinion plays a limited role in governments' decision to abolish the death penalty. Zimring and Hawkins (1986) explain that ...... However, scholars warn against generalizing this trend onto the United States, as American democracy is riddled with idiosyncratic features that alter the way public opinion interacts with governance and governmental decision-making. These features include (but are not limited to) popularly elected prosecutors and judges, the frequent use of referendums and ballot initiatives, weak political parties and 'intermediary institutions,' primary elections, and a federal system of government that delegates substantial power to state and local governments (Garland, 2010; Steiker, 2006). These features heighten the salience of public opinion in governmental decision-making. Therefore, there is reason to believe American lawmakers might be responsive to public opinion when it comes to capital punishment and executions. Guided by this literature, I present my first hypothesis:

**H1: A high level of public support for capital punishment will increase the number of executions carried out in a given year.**

For H1, my explanatory variable is the level of public support for capital punishment among surveyed Americans. This variable is measured in percentage points. Specifically, this variable is measured as the share of survey respondents who answered "yes" to the the following survey question: "Are you in favor of the death penalty for persons convicted of murder?" My response variable is the number of executions carried out in the United States in a given year. This variable is measured simply in the raw number of individuals executed. H1 would be supported by the observation that a greater number of executions are carried out when public support for capital punishment is high compared to when public support for capital punishment is low. The following two observations would disprove H1: (1) there is no correlation between the level of public support for capital punishment and the number of executions, or (2) fewer executions are carried out when public support for capital punishment is high compared to when public support is low.

Next, ___ [crime rate] As such, I present a second hypothesis:

**H2: As the national homicide rate increases, the number of executions will increase.**

For H2, my explanatory variable is the national homicide rate, which is measured as ___. My response variable, once again, is the number of executions carried out in the United States in a given year. H2 would be supported by the observation that more executions are carried out during periods of high rates of homicide compared to periods of low rates of homicide. There are two main observations that would disprove this hypothesis: (1) there is no correlation between the homicide rate and the number of executions, or (2) the number of executions tends to be higher during periods of low rates of homicide compared to periods of high rates of homicide.

Finally, Steiker & Steiker (2012) explain that, upon adjudication of Gregg v. Georgia, the US became the "first and only one of its peer nations to move not from formal retention of the death penalty to abolition, but rather from retention to regulation." This procedural regulation, according to Steiker & Steiker (1995), has had the “effect of legitimating the use of capital punishment as a penal sanction in the eyes of actors within the criminal justice system and the public at large." In other words, any alteration to, or regulation of, the institution of capital punishment that gives it the appearance of being less "cruel and unusual" further legitimizes executions and entrenches the institution of capital punishment legally and socially. Guided by the research of Steiker & Steiker, I have formulated a third (and final) hypothesis:

**H3: The number of executions will increase as the number of death row exonerations increases.**

For H3, my explanatory variable is the number of death row 


## Data & Research Design
My data does not come from a single source, as I was unable to find a dataset that contained all (or even most) of the variables I was interested in analyzing. Instead, I compiled data from five different datasets into a single excel spreadsheet, which I then converted to a .csv file. My data on the number of annual executions comes from the U.S. Department of Justice's Bureau of Justice Statistics (BJS). The BJS database contained data for the period 1977-2022, in a dataset entitled "Prisoners Executed." Due to the Furman v. Georgia ruling, no executions were carried out between 1972-1976, so I was able to _____ in my Excel spreadsheet. For data on the annual national homicide rate, I used 


The BJS website no longer publicly displays homicide rdata for years prior to 1985, so in order to find this information, I searched the Internet Archive's Wayback Machine. There, I located an archived dataset published by 


For data on the number of death row exonerations, I used the Death Penalty Information Center's


Finally, for data on public opinion/public support for capital punishment

| Name                   | Description                               
|------------------------|--------------------------------------------------------------------
| `Year`                 | Year
| `Percent_Support`      | Percentage of surveyed Americans who favor the death penalty as a punishment for murder
| `Number_Executions`    | Number of prisoners executed in the United States
| `Number_Exonerations`  | Number of death row exonerations
| `Homicide_Rate`        | National homicide rate (measured in homicides per 100,000 residents)
| `Fin_Dissatisfaction`  | Percentage of surveyed Americans who are dissatified with their overall financial situation

I included a few variables that I do not use in this report, but that may yield fruitful inside if analyzed in future research. The data for each of these variables also came from the General Social Survey dataset. Below are the variables and a brief description of each:

| Name                          | Description                               
|-------------------------------|--------------------------------------------------------------------
| `Pro_Prayer_Schools`          | Percentage of surveyed Americans who support Christian prayer in public schools
| `Ban_IRM`                     | Percentage of surveyed Americans who support a ban on interracial marriage
| `Insufficient_Spending_CR`    | Percentage of surveyed Americans who believe the government spends too little money addressing the issue of the rising crime rate
| `Distrust_SC`                 | Percentage of surveyed Americans who distrust the Supreme Court
| `Distrust_Congress`           | Percentage of surveyed Americans who distrust Congress


I then loaded this file into RStudio, named it "deathpenalty," and presented my dataset in a table. 

```{r}
library(tidyverse)
library(dplyr)
library(broom)
library(infer)
library(modelsummary)

deathpenalty <- read_csv("deathpenalty.csv")
knitr::kable(deathpenalty)

```

To visualize key trends in my dataset, I first plotted the number of executions per year, both as a line graph and as a scatter plot with a smoothed trend line.

```{r}
dp_plot_1 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Number_Executions)) + 
  geom_line(color = "red") + 
  labs(x = "Year", 
       y = "Number of Executions", 
       title = "Number of Executions in the United States per Year", 
       caption = "Source: US Department of Justice")

dp_plot_2 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Number_Executions)) + 
  geom_point() + geom_smooth(color = "red") + 
  labs(x = "Year", 
       y = "Number of Executions", 
       title = "Number of Executions in the United States per Year", 
       caption = "Source: US Department of Justice")

dp_plot_1
dp_plot_2
```

Then, I plotted the key explanatory variable for each of my three hypotheses over time as both a line graph and a scatter plot with a smoothed trend line. These variables are percentage of Americans in favor of capital punishment, national homicide rate, and number of exonerations, respectively.

```{r}
dp_plot_3 <- ggplot(data = deathpenalty[complete.cases(deathpenalty$Percent_Support),],
  mapping = aes(x = Year, y = Percent_Support)) + 
  geom_line() +
  labs(x = "Year", 
       y = "Percent in Favor", 
       title = "% of Americans in Favor of Capital Punishment for Homicide by Year", 
       caption = "Source: General Social Survey") 
  

dp_plot_4 <- ggplot(data = deathpenalty,  
  mapping = aes(x = Year, y = Percent_Support)) + 
  geom_point() + 
  geom_smooth(color = "black") + 
  labs(x = "Year", 
       y = "Percent in Favor", 
       title = "% of Americans in Favor of Capital Punishment for Homicide by Year", 
       caption = "Source: General Social Survey")

dp_plot_5 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Homicide_Rate)) + 
  geom_line(color = "darkorchid4") + 
  labs(x = "Year", 
       y = "Homicide Rate", 
       title = "Homicide Rate in the United States per Year", 
       caption = "Source: Federal Bureau of Investigations")

dp_plot_6 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Homicide_Rate)) + 
  geom_point() +
  geom_smooth(color = "darkorchid4") + 
  labs(x = "Year", 
       y = "Homicide Rate", 
       title = "Homicide Rate in the United States per Year", 
       caption = "Source: Federal Bureau of Investigations")

dp_plot_7 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Number_Exonerations)) + 
  geom_line(color = "forestgreen") + 
  labs(x = "Year", 
       y = "Number of Exonerations", 
       title = "Number of Exonerations in the United States per Year", 
       caption = "Source: US Department of Justice")

dp_plot_8 <- ggplot(data = deathpenalty,
  mapping = aes(x = Year, y = Number_Exonerations)) + 
  geom_point() + 
  geom_smooth(color = "forestgreen") + 
  labs(x = "Year", 
       y = "Number of Exonerations", 
       title = "Number of Exonerations in the United States per Year", 
       caption = "Source: US Department of Justice")

dp_plot_3
dp_plot_4
dp_plot_5
dp_plot_6
dp_plot_7
dp_plot_8
```


## Results

**Bivariate Regressions**

***Testing Hypothesis 1***
```{r}
varnames1 <- c("(Intercept)" = "Intercept",
               "Percent_Support" = "Public Support")

dp_lm_1 <- lm(Number_Executions ~ Percent_Support, 
   data = deathpenalty)
summary(dp_lm_1)
modelsummary::modelsummary(list(Regression = dp_lm_1),
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"),
                           coef_map = varnames1)

dp_plot_9 <- ggplot(data = deathpenalty,
mapping = aes(x = Percent_Support, y = Number_Executions)) +
geom_point() +
geom_smooth(color = "black") +
  labs(x = "Percent in Favor", y = "Number of Executions", 
       title = "Number of Executions vs. Support for Capital Punishment", 
       caption = "Sources: General Social Survey & US Department of Justice")

dp_plot_9
```
According to this regression, the intercept is 12.410, which means that — all else being constant — when 0% of Americans support capital punishment, 12.410 individuals will be executed annually. The coefficient for public support  is 0.183, which means that — all else being held constant — for each unit (percentage point) of increase in public suoport for capital punishment, the number of executions increases by .183 units (persons executed). However, this relationship is not statistically significant, as the p-value is well above 0.05. Therefore, I am not able to reject the null hypothesis that public support  is not correlated with number of executions. In addition, these results should not be interpreted causally, as this regression does not take any or all possible confounding variables into account.

***Testing Hypothesis 2***
```{r}
varnames2 <- c("(Intercept)" = "Intercept",
               "Homicide_Rate" = "Homicide Rate")

lm_dp_2 <- lm(Number_Executions ~ Homicide_Rate,
              data = deathpenalty)
summary(lm_dp_2)
modelsummary::modelsummary(list(Regression = lm_dp_2),
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"),
                           coef_map = varnames2)

dp_plot_10 <- ggplot(data = deathpenalty,
mapping = aes(x = Homicide_Rate, y = Number_Executions)) +
geom_point() +
geom_smooth(color = "darkorchid4") +
  labs(x = "Homicide Rate", y = "Number of Executions", 
       title = "Number of Executions vs. Homicide Rate", 
       caption = "Sources: General Social Survey & Federal Bureau of Investigations")

dp_plot_10

```
According to this regression, the intercept is 91.697, which means that — all else being constant — when the homicide rate is 0, 91.697 individuals will be executed annually. The coefficient for homicide rate is -8.504, which means that — all else being held constant — a one-unit (homicide per 100,000 individuals) increase in homicide rate would be associated with a 8.504-unit (persons) decrease in number of executions. This relationship is  statistically significant, as the p-value is well below 0.05. Therefore, I can reject the null hypothesis that homicide rate is not correlated with number of executions. That said, these results should not be interpreted causally, as this regression does not take any or all possible confounding variables into account. 

***Testing Hypothesis 3***
```{r}
varnames3 <- c("(Intercept)" = "Intercept",
               "Number_Exonerations" = "Number of Exonerations")

lm_dp_3 <- lm(Number_Executions ~ Number_Exonerations,
              data = deathpenalty) 
summary(lm_dp_3)
modelsummary::modelsummary(list(Regression = lm_dp_3),
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"),
                           coef_map = varnames3)

dp_plot_11 <- ggplot(data = deathpenalty,
mapping = aes(x = Number_Exonerations, y = Number_Executions)) +
geom_point() +
geom_smooth(color = "forestgreen") +
  labs(x = "Number of Exonerations", y = "Number of Executions", 
       title = "Number of Executions vs. Number of Exonerations", 
       caption = "Sources: Death Penalty Information Center & US Department of Justice")

dp_plot_11
```
According to this regression, the intercept is 16.035, which means that — all else being constant — when there are 0 exonerations, 16.035 individuals will be executed annually. The coefficient for number of exonerations is 3.939, which means that — all else being held constant — a one-unit (person) increase in the number of exonerations would be associated with a 3.939-unit (person) increase in the number of executions. This relationship is  statistically significant, as the p-value is below 0.05. Therefore, I am able to reject the null hypothesis that number of exonerations is not correlated with number of executions. However, these results should not be interpreted causally, as this regression does not take any or all possible confounding variables into account.

**Multiple Regression**
***Further Testing of Hypothesis 3***
```{r}
varnames4 <- c("(Intercept)" = "Intercept",
               "Number_Exonerations" = "Number of Exonerations",
               "Percent_Support" = "Public Support",
               "Homicide_Rate" = "Homicide Rate")

lm_dp_4 <- lm(Number_Executions ~ Number_Exonerations + Homicide_Rate + Percent_Support,
              data = deathpenalty) 
summary(lm_dp_4)
modelsummary::modelsummary(list(Regression = lm_dp_4),
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"),
                           coef_map = varnames4)
```
According to this regression, the intercept is -7.144, which means that, when level of public support, homicide rate, and number of exonerations are all held at 0, -7.144 individuals would be executed annually. The coefficient for number of exxonerations is 2.949, which means that—when level of public support and  homicide rate are held constant—a one-unit (person) increase in number of exonerations would be associated with a 2.949-unit (person) increase in the number of executions. As was the case in the bivariate regression, this relationship is statistically significant, as the p-value is under 0.05. However, both the coefficient and the standard error for the number of exonerations is lower in this multiple regression than they were in the bivariate regression.

The outputs of this multiple regression are particularly interesting when analyzed in light of the outputs of the three preceding bivariate regressions. Perhaps most notable is the fact that the relationship between public support and number of executions is NOT statistically significant when calculated in a bivariate regression, but IS statistically significant when the homicide rate and the number of exonerations are held constant in a multiple regression.




## Further/Secondary Analysis

For a bit of secondary analysis, I decided to investigate potential factors or conditions that impact the level of public support for capital punishment. First, to explore the relationship between the homicide rate and public support for the death penalty, I ran a regression using "Percent_Support" as the response variable and "Homicide_Rate" as the explanatory variable, and plotted these two variables on a scatter plot with a smoothed trend line. 


```{r}
varnames5 <- c("(Intercept)" = "Intercept",
               "Homicide_Rate" = "Homicide Rate")

lm_support_1 <- lm(Percent_Support ~ Homicide_Rate,
              data = deathpenalty) 
summary(lm_support_1)
modelsummary::modelsummary(list(Regression = lm_support_1),
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"),
                           coef_map = varnames5)

support_plot_1 <- ggplot(data = deathpenalty,
  mapping = aes(x = Homicide_Rate, y = Percent_Support)) + 
  geom_point() + geom_smooth(color = "darkorchid4") + 
  labs(x = "Homicide Rate", 
       y = "Percent in Favor", 
       title = "Percent in Favor vs. Homicide Rate", 
       caption = "Sources: General Social Survey & US Department of Justice")

support_plot_1
```
According to this regression, the intercept is 56.884, which means that — all else being constant — when the homicide rate is 0,  public support for capital punishment is 56.884%. The coefficient for homicide rate is 1.311, which means that — all else being held constant — for each unit (FIGURE OUT UNIT) the homicide rate increases, public support for capital punishment increases by 1.311 units (percentage points). This relationship is statistically significant, as the p-value is under 0.05. Therefore, I am able to reject the null hypothesis that the homicide rate is not correlated with public support for capital punishment. That said, these results should not be interpreted causally, as this regression does not take any or all possible confounding variables into account.


Then, to explore the relationship between the rate of financial dissatisfaction and public support for the death penalty, I ran a regression using "Percent_Support" as the response variable and "Fin_Dissatisfaction" as the explanatory variable. I also plotted these two variables on a scatter plot with a smoothed trend line. 

```{r}
varnames6 <- c("(Intercept)" = "Intercept",
               "Fin_Dissatisfaction" = "Financial Dissatisfaction")

lm_support_2 <- lm(Percent_Support ~ Fin_Dissatisfaction,
              data = deathpenalty) 
summary(lm_support_2)
modelsummary::modelsummary(list(Regression = lm_support_2),
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"),
                           coef_map = varnames6)

support_plot_2 <- ggplot(data = deathpenalty,
  mapping = aes(x = Fin_Dissatisfaction, y = Percent_Support)) + 
  geom_point() + geom_smooth(color = "blue") + 
  labs(x = "Financial Dissatisfaction", 
       y = "Percent in Favor", 
       title = "Percent in Favor vs. Financial Dissatisfaction", 
       caption = "Sources: General Social Survey & US Department of Justice")

support_plot_2
```
According to this regression, the intercept is 61.167, which means that — all else being constant — when the rate of financial dissatisfaction rate is 0,  public support for capital punishment is 61.167%. The coefficient for financial dissatisfaction is 0.223, which means that — all else being held constant — for each unit (percentage point) that financial dissatisfaction increases, public support for capital punishment increases by 0.223 units (percentage points). However, this relationship is NOT statistically significant, as the p-value is well above 0.05. Therefore, I fail to reject the null hypothesis that financial dissatisfaction has no correlation/relationship to the level of public support for capital punishment. Additionally, I cannot infer causality from this analysis, as this regression does not hold any or all potential confounding variables constant.


Finally, I ran a multiple regression using Percent_Support as the response variable and "Homicide_Rate" and "Fin_Dissatisfaction" as the explanatory variables.

```{r}
lm_support_3 <- lm(Percent_Support ~ Homicide_Rate + Fin_Dissatisfaction,
              data = deathpenalty) 

summary(lm_support_3)
modelsummary::modelsummary(list(Regression = lm_support_3),
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```
In this multiple regression, the relationship between the homicide rate and public support remains statistically significant and the relationship between the rate of financial dissatisfaction and public support remains not statistically significant. However, compared to the __ regressions, the coefficient and the standard error for homicide rate increased slightly, and its p-value decreased. ___

## Conclusion
Ultimately, my analyses provide some support for H1 (). The bivariate regression returned findings that aligned with my hypothesis but were not statistically signficant. However, a multiple regression that __ homicide rate and number of exonerations, returned findings that aligned with my hypothesis and were statistically significant. From this, I can conclude that ___. However, from this alone I cannot infer causality, as .....

For H2, my findings show a statistically significant relationship between the homicide rate and the number of executions. However, contrary to my hypothesis, this relationship/correlation is negative. In other words, as the homicide rate increases, the number of executions decrease, and as the homicide rate decreases, the number of executions increase. Therefore, my results run counter to H2. This finding was particularly surprising to me and certainly warrants further research. One potential explanation may be reverse causality. Though still widely debated, some scholars maintain that capital punishment/executions deter violent crime. Further analyses, then, should investigate the relationship between these two variables using the number of executions as the explanatory variable and the homicide rate as the response variable. 

For H3, my findings show a statistically significant relationship between the number of exonerations and the number of executions. As predicted, I found that the number of executions increase as the number of exonerations increase. As some scholars suggest, this may be attributed to the legitimating/stabilizing effect that procedural regulation has on the institution of capital punishment. Exonerations, therefore—by demonstrating that mistakes/wrongful convictions can be rectified—may work to legitimate executions and entrench the institution of capital punishment. Governmental actors may be more likely to carry out excecutions (and may receive less push-back for doing so) if they have greater confidence that the wrongfully convicted will be/have been exonerated.


One limitation of my analysis is that it does not break down my data on a state-by-state basis. Since states have original jurisdiction over their criminal justice policies and carry out their own executions, attaining a full understanding of the link between explanatory variables and the number of executions would require a state-by-state breakdown, along with this national-level analysis. Hypothetically, a state without the death penalty could have a disproportionately high homicide rate. If this were the case, it may help to explain the surprising findings of my tests of Hypothesis 2. Unfortunately, to my best knowledge, state-by-state public opinion data on capital punishment is not publicly available. 

Another limitation is that my analyses do not consider the number of death sentences given each year or the number of people on death row each year. If I had unlimited time, I'd like to repeat these regressions using the proportion of death row inmates executed each year along with the raw-number strategy I used in this study. I'd also analyze the relationship between death sentences per year and the other variables I analyzed in this study.


## Works Cited
__Data:__

Bureau of Justice Statistics. (2011). "Homicide Trends In The United States." U.S. Department of
 	Justice. Data accessed using the Internet Archive’s Wayback Machine,
 	https://web.archive.org/web/20110727134213/http://bjs.ojp.usdoj.gov/content/homicide/tables/totalstab.cfm

Crime Data Explorer. (2023). "Trend of Homicide from 1985 to 2022." Federal Bureau of Investigations.
  https://cde.ucr.cjis.gov/LATEST/webapp/#/pages/explorer/crime/crime-trend 

Davern, Michael et. al., National Opinion Research Center at the University of Chicago. (2023).
  "General Social Survey 1972-2022." Sponsored by National Science Foundation. Data
  accessed from the GSS Data Explorer website at gssdataexplorer.norc.org.
  
Death Penalty Information Center. (2023). Innocence Database.
  https://deathpenaltyinfo.org/database/innocence

Snell, Tracy L., U.S. Department of Justice. (2023). "Prisoners Executed." Bureau of Justice
  Statistics. https://bjs.ojp.gov/library/publications/prisoners-executed.
  

  
  
__Academic Literature:__

Garland, David. "Peculiar Institution: America’s Death Penalty in an Age of Abolition."
Cambridge, MA: Belknap Press of Harvard University Press, 2010.

Killingley, Julian. "Constraining America's Death Penalty: The Eighth Amendment and Excessive
Punishment." in "Against the Death Penalty: International Initiatives and Implications." ed. 
John Yorke. Ashgate Publishing, 2008.

Steiker, Carol S. “Capital Punishment and American Exceptionalism,” in "American Exceptionalism and Human Rights," ed. Michael Ignatieff (Princeton University Press, 2006).

Steiker, Carol S., and Jordan M. Steiker. "Courting Death: The Supreme Court and Capital
  Punishment." Cambridge, MA: The Belknap Press of Harvard University Press, 2016.

Steiker, Carol S., and Jordan M. Steiker. “Entrenchment and/or Destabilization? Reflections on
  (Another) Two Decades of Constitutional Regulation of Capital Punishment.” Law and
  Inequality: A Journal of Theory and Practice 30, no. 2 (August 10, 2012).

Steiker, Carol S., and Jordan M. Steiker. “Sober Second Thoughts: Reflections on Two Decades
  of Constitutional Regulation of Capital Punishment.” Harvard Law Review 109, no. 2 (1995): 355–438.
  
Zimring, Frankin E. and Gordon Hawkins. "Capital Punishment and the American Agenda."
Cambridge University Press, 1986.


